Cognate beginnings to bilingual lexical acquisition

### 2023-05-26 manuscript/manuscript.qmd
Fix typos

abstract: |
    Language non-selectivity is prominent feature of bilingual lexical processing. An instance of such non-selectivity is embodied by cognateness, which recent studies have suggested to facilitate vocabulary acquisition in bilingual toddlers, who show larger vocabulary sizes when their languages share many cognates, and to be acquired at earlier ages than non-cognates. The specific mechanisms behind such facilitation are unclear. We present an account of early bilingual vocabulary growth in which cognateness interacts with lexical frequency and language exposure to facilitate the word acquisition. We evaluated this model against vocabulary data from 436 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the toddler. We found cognateness facilitation effect only in low-frequency words. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

One way typological distance may impact bilingual trajectories of language acquisition is the presence (or absence) of *cognates* in their linguistic input. Quite often, bilingual children encounter more than one label for the same referent: one in each language. For instance, an English-Spanish bilingual might hear both *dog* and *perro* in the presence of a dog We will refer to such cross-language synonyms as *translation equivalents*.^[Throughout the present study, we will make the assumption that bilinguals form a common semantic representations for each translation equivalent (e.g., the concept 'dog' is shared for the English and Spanish words *dog* and *perro*). The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight parts of the body (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.].

Sometimes, the word-forms involved in a translation equivalent may be phonologically similar. Such form-similar translation equivalents are referred to as cognates. While the previously mentioned translation pair *dog* and *perro* are non-cognates (both word-forms hardly share any phonemes), the translation pair *cat* and *gato* can be considered cognates. The amount of cognates shared by a pair of languages is dependent on their typological distance. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like English and Spanish (Romance and Germanic languages, respectively). Therefore, bilingual children learning two typologically close languages may be exposed to a larger amount of cognates than those learning two typologically more distant languages. For instance, in the presence of a door, a Catalan-Spanish bilingual might be exposed to the words *porta* and *puerta* (cognate translation equivalent with the common meaning 'door'), whereas an English-Spanish bilingual might hear *door* and *puerta* (non-cognate translation equivalent).

@@

The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/\textipa{"kæt}/-/\textipa{"kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /\textipa{æ}/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /\textipa{"kæt}/-/\textipa{"Ga.ta}/) would score three, as it requires at least three editions: substituting /\textipa{k}/ for /\textipa{G}/, substituting /\textipa{\ae}/ for /\textipa{a}/ and adding another /\textipa{a}/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /\textipa{"kæt}/-/\textipa{"kAt}/, and $3/4 = 0.75$ for /\textipa{"kæt}/-/\textipa{"Ga.ta}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes. When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity, the amount of co-activate of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. This facilitation effect may not be equivalent for /\textipa{"gat}-\textipa{"ga.to}/. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show a stronger cognateness facilitation effect than words in Catalan.

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. We illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect through parallel activation, in which both word-forms co-activate each other accumulating learning instances faster, and (2) no cognateness facilitation, for which both word-forms accumulate learning instances independently.

@@

plot_h0 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H0: no parallel activation",
         linetype = "Hypothesis") +
    scale_x_continuous(breaks = seq(min(eli_df$age),
                                    max(eli_df$age), 4)) +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1.5, "cm"),
          legend.margin = margin(c(0, 0, 0, 0)),
          legend.justification = c(0, 1),
          legend.title = element_text(size = 9),
          legend.text= element_text(size = 8))

@@

aoa_df <- generate_aoa(eli_df, threshold = 250) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H1")

plot_h1 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H1: parallel activation",
         linetype = "Hypothesis") +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    scale_x_continuous(breaks = seq(min(eli_df$age),
                                    max(eli_df$age), 4)) +
    theme(legend.position = "none")

plot_h0 + plot_h1 +
    plot_layout(ncol = 1, guides = "collect") &
    plot_annotation(tag_levels = "A") +
    scale_colour_manual(values = clrs[c(1, 4)]) &
    scale_fill_manual(values = clrs[c(1, 4)]) &
    scale_y_continuous(labels = function(x) format(x, big.mark = ",")) &
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) &
    theme_ggdist() &
    theme(legend.position = "top", 
          legend.text = element_text(size = 10),
          legend.key.width = unit(2, "cm"),
                    legend.key.height = unit(0.5, "cm"),
          strip.background = element_rect(fill = "grey90", colour = "grey90"),
          strip.text = ggtext::element_markdown(size = 10),
          panel.grid.major.y = element_line(colour = "grey",
                                            linetype = "dotted"),
          plot.background = element_rect(fill = "white",
                                         colour = NA))

@@

We tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a word *only* once its translation has been already acquired. We collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian explanatory item response theory  [IRT, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word-form, while adjusting for the the number of phonemes in the word-form (an indicator of word acquisition difficulty).

@@

*Cognateness* was defined as a continuous measure of phonological similarity between translation equivalents. We quantified the phonological similarity between each pair of Catalan-Spanish word-forms by computing the inverse of the normalised Levenshtein distance between their X-XSAMPA phonetic transcriptions [@levenshtein1966binary; @wells1995computercoding]. This score is computed by counting the number of insertions, deletions and replacements needed by both transcriptions to become identical, dividing the resulting value by the length of the longest transcription, and finally subtracting this value from one. This results in a proportion that indicates how much the two phonetic transcriptions of the translation equivalent are similar to each other, ranging from zero (no similarity at all) to one (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. For instance, this measure of phonological similarity returns 0% for *ocell* /`uceL`/ and *pájaro* /`paxa4o`/, Catalan and Spanish for 'bird', and returns 50% for *lluna* /`Lun5`/ and *luna* /`luna`/, Catalan and Spanish for 'moon' (see Appendix A for more details).

@@

r_version <- glue::glue(r_info$R.version$major, 
                        r_info$R.version$minor, 
                        .sep = ".")
```

From the responses to the `r format(nrow(n_items), big.mark = ",")` items included in the questionnaire, we restricted the analyses to responses to nouns (`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded) [@fourtassi2020growth]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to words `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, mixed-effects ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (composite measure of Zipf-transformed lexical frequency and language degree of exposure), and *Cognateness* (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [see @eq-linear, @barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

>>>>

abstract: |
    Language non-selectivity is a prominent feature of bilingual lexical processing. An instance of such non-selectivity is embodied by cognateness. Recent studies have pointed to cognateness as a facilitator of vocabulary growth, and to cognates being acquired at earlier ages than non-cognates. The specific mechanisms behind such facilitation are unclear. Building on accumulator models of language acquisition, we suggest that cognateness increases the number of cumulative learning instances that a bilingual child encounters, thanks to cross-language activation. In line with previous studies, this model predicted a cognate facilitation, but only for words to which children are exposed less often. We tested these predictions against vocabulary data from 366 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the participant. We found a cognateness facilitation effect only in low-frequency words or words from the language of lower exposure. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

One way typological distance may impact bilingual trajectories of language acquisition is the presence (or absence) of *cognates* in their linguistic input. Quite often, bilingual children encounter more than one label for the same referent: one in each language. For instance, an English-Spanish bilingual might hear both *dog* and *perro* in the presence of a dog We will refer to such cross-language synonyms as *translation equivalents*.^[Throughout the present study, we make the assumption that bilinguals form a common semantic representations for each translation equivalent (e.g., the concept 'dog' is shared for the English and Spanish words *dog* and *perro*). The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight body parts (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.]. Sometimes, the word-forms involved in a translation equivalent may be phonologically similar. Such form-similar translation equivalents are referred to as *cognates*. While the previously mentioned translation pair *dog* and *perro* are non-cognates (both word-forms hardly share any phonemes), the translation pair *cat* and *gato* can be considered cognates. The amount of cognates shared by a pair of languages is dependent on their typological distance. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like English and Spanish (Romance and Germanic languages, respectively). Therefore, bilingual children learning two typologically close languages may be exposed to a larger amount of cognates than those learning two typologically more distant languages. For instance, in the presence of a door, a Catalan-Spanish bilingual might be exposed to the words *porta* and *puerta* (cognate translation equivalent with the common meaning 'door'), whereas an English-Spanish bilingual might hear *door* and *puerta* (non-cognate translation equivalent).

@@

The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/\textipa{"kæt}-\textipa{"kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /\textipa{æ}-\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /\textipa{"kæt}-\textipa{"Ga.ta}/) would score three, as it requires at least three editions: substituting /\textipa{k}/ for /\textipa{G}/, substituting /\textipa{\ae}/ for /\textipa{a}/ and adding another /\textipa{a}/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /\textipa{"kæt}-\textipa{"kAt}/, and $3/4 = 0.75$ for /\textipa{"kæt}-\textipa{"Ga.ta}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes (0.75 Levenshtein similarity). When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity (0% Levenshtein similarity), the amount of co-activate of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. This facilitation effect may not be equivalent for /\textipa{"gat}-\textipa{"ga.to}/. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show an even stronger cognateness facilitation effect than words in Catalan.

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 60% of the time to Catalan, and 40% of the time to Spanish. We also assume that the child is exposed to either word around 50 times per month, following a $\text{Poisson}(\lambda = 50)$ distribution. We illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{"gat}/ or /\textipa{"gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect through parallel activation, in which both word-forms co-activate each other accumulating learning instances faster, and (2) no cognateness facilitation, for which both word-forms accumulate learning instances independently.

@@

plot_h0 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             size = 2,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H0: no parallel activation",
         linetype = "Hypothesis") +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1.5, "cm"),
          legend.margin = margin(c(0, 0, 0, 0)),
          legend.justification = c(0, 1),
          legend.title = element_text(size = 9),
          legend.text= element_text(size = 8))

@@

aoa_df <- generate_aoa(eli_df, threshold = threshold) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H1")

plot_h1 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             size = 2,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H1: parallel activation",
         linetype = "Hypothesis") +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    theme(legend.position = "none")

plot_h0 + plot_h1 +
    plot_layout(ncol = 1, guides = "collect") &
    plot_annotation(tag_levels = "A") +
    scale_colour_manual(values = clrs[c(1, 4)]) &
    scale_fill_manual(values = clrs[c(1, 4)]) &
    scale_y_continuous(labels = function(x) format(x, big.mark = ","),
                       limits = c(0, 1500)) &
    scale_x_continuous(labels = function(x) format(x, big.mark = ","),
                       limits = c(0, 50)) &
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) &
    theme_ggdist() &
    theme(legend.position = "top", 
          legend.text = element_text(size = 10),
          legend.key.width = unit(2, "cm"),
          legend.key.height = unit(0.5, "cm"),
          strip.background = element_rect(fill = "grey90", colour = "grey90"),
          strip.text = ggtext::element_markdown(size = 10),
          panel.grid.major.y = element_line(colour = "grey",
                                            linetype = "dotted"),
          plot.background = element_rect(fill = "white",
                                         colour = NA))

@@

We tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word. We collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian explanatory item response theory  [IRT, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word-form, while adjusting for the the number of phonemes in the word-form (an indicator of word acquisition difficulty).

@@

*Cognateness* was defined as a continuous measure of phonological similarity between translation equivalents. We quantified the phonological similarity between each pair of Catalan-Spanish word-forms by computing the inverse of the normalised Levenshtein distance between their X-SAMPA phonetic transcriptions [@levenshtein1966binary; @wells1995computercoding]. This score is computed by counting the number of insertions, deletions and replacements needed by both transcriptions to become identical, dividing the resulting value by the length of the longest transcription, and finally subtracting this value from one. This results in a proportion that indicates how much the two phonetic transcriptions of the translation equivalent are similar to each other, ranging from zero (no similarity at all) to one (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. For instance, this measure of phonological similarity returns 0% for *ocell* /`uceL`/ and *pájaro* /`paxa4o`/, Catalan and Spanish for 'bird', and returns 50% for *lluna* /`Lun5`/ and *luna* /`luna`/, Catalan and Spanish for 'moon' (see Appendix A for more details).

@@

r_version <- glue::glue(r_info$R.version$major, 
                        r_info$R.version$minor, 
                        .sep = ".")
stan_version <- cmdstanr::cmdstan_version()

We collected `r format(nrow(n_items), big.mark = ",")` items responses. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, @fourtassi2020growth]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to words `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (composite measure of Zipf-transformed lexical frequency and language degree of exposure), and *Cognateness* (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [see @eq-linear, @barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

$$
\begin{aligned}
&\textbf{Response}~(k)~\textbf{to word}~i~ \textbf{by participant}~j \\
\text{Response}_{ij} &\sim \text{Cumulative logit}(\theta_{k_{ij}}), \\ 
&\text{where}~k \in \{\text{No} \rightarrow \text{Understands}, \text{Understands} \rightarrow \text{Understands and Says}\}\\
&\textbf{Distribution parameters} \\
\theta_{k_{ij}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + \\
&(\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + 
(\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) \\
&\textbf{Prior} \\
\beta_{0_{k}} &\sim \mathcal{N}(-0.25, 0.5); ~\beta_{1-5} \sim \mathcal{N}(0, 1) \\
\sigma_{u_{0-8}, w_{0-3}} &\sim \mathcal{N_{+}}(1, 0.25); ~\rho_{u_{0-8}, w_{0-3}} \sim \text{LKJcorr}(2)
\end{aligned}
$$ {#eq-linear}
